Graphical Tests for Power Comparison of Competing Designs

### 2012-03-31 intro.tex
all citet are now cite

% !TEX root = comparison.tex
%% \section{Introduction} %for journal use above \firstsection{..} instead
%We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
So, you've just spent countless hours burning the late night candle
wax and discovered several astonishing features in the data. Your
exploratory graphics are quick to produce, interactive, flexible, and
fabulous for discovery, but now you need to present the information to
your boss, the co-workers, a class of students, or get it published in
that top-rated journal. Those exploratory graphics are unlikely to
communicate the discovered information to your audience efficiently,
or elegantly.  To decide on the best way to present the specific
information there can be many types of plots to choose from. The
decision process is assisted by past experience, personal aesthetic
preferences, the characteristics of the audience, and a good working
knowledge of general perceptual strengths and weaknesses of particular
displays \citep{cleveland:1984,heer:2010}. Perceptual studies are
thinly spread for data visualization and will probably
only cover very broad perceptual principles, primarily because they
tend to be based solely on simulated data. In order to study
perceptual principles in a controlled experimental setting simulated
data is unavoidable. This provides principles that may not apply
closely enough to a specific analysis task.  Faced with design
decisions for a very specific task requires a closer level of
detail. Focus groups \citep{Mazza07} can find small problems with
designs and move us closer to a final handful of possible
designs. Usability studies and case studies
(e.g. \citet{Plaisant:2004}, \citet{Shneiderman:2006}) typically focus
on the use of particular software for solving data exploration tasks.
These approaches lack broad audience validity and
objectiveness. Several papers indicate continuing issues arising with,
and suggestions for improving, evaluation of information
visualizations
\citep{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}.

@@

Here we suggest a new approach using recent research on visual inference \citep{buja:2009,wickham:2010}. In visual inference, data plots are considered to be test statistics, and these are compared with plots of data generated from a null hypothesis using a lineup. The null hypothesis underlying a lineup is that there is no real structure visible in the data plot, that any patterns seen are consistent with randomness, or from a known model. If the null hypothesis is true then the plot of the data will not be distinguishable form the plots of null data. A lineup is similar to the police lineup, and this is where the name comes from. The suspect (the plot of the data) is placed randomly in a field of plots of null data, and an impartial viewer is asked to identify the plot that is the most different. If the viewer identifies the suspect it lends statistical significance to the conclusion that the data is not consistent with the null data. Figure \ref{lineup} has an example of a lineup. Overlaid density plots are used to display the distribution of two groups (blue, red) of data. We are interested to know whether the data has a significant difference between the centers of the two groups. Plot numbered 13 contains the data. Did you pick that one? In this plot the blue group has a flat density whereas the density of red group is more peaked and concentrated to the left of the blue.

@@

The previous work \citet{buja:2009,wickham:2010} describes the use of lineups for evaluating discoveries made in exploratory analysis. Here we describe the use of lineups for evaluating plot design for communicating findings visually. Different lineups are created, using the same data, same null plots and positions within the lineup, but different plot designs. Comparing how long and how accurately the viewers take to report the feature of interest will assess which design is better for the task. To investigate the benefit of lineups for plot design evaluation we have conducted an experiment that also contains simulated data, where the data is constructed to match and expand on the features of interest in the data. For example, in a simple situation, where the discovered feature of interest is a difference between the centers of two distributions, simulated data would be constructed roughly matching the distribution of the data that varies the distance between centers, and also spread and sample size. Lineups of these simulated data (embedded with simulated null data plots) are generated and evaluated along with the lineups of data. The simulation study provides a backdrop for the real data, which enables the broad applicability to be studied and a gauge for the strength of the pattern in the data.

@@

Lineups provide a way to statistically quantify the significance of the finding \citet{majumder:2011}: think of the lineup as a set of $m$ plots, one of which is the data. (In Figure \ref{lineup}, $m=20$.)  The probability that an observer picks the data plot just by chance, i.e. when it is really *not* different from the other null plots, is $1/m$.  If an observer is able to identify the data plot from the lineup, we reject the null hypothesis.  This sets our Type I error rate, $\alpha$, at a level of $1/m$.

@@

When comparing different tests of the same quantity, we consider that
test better if it has greater {\it power}.  The power of a test is the
probability to reject the null hypothesis, irrespective of whether it
is true or false -- in a lineup this is the probability that an
observer identifies the plot of the real data.  Analytically, power is
usually difficult to calculate because it is requires specific use of
an alternative hypothesis to calculate the probability.  Work in
\citet{niladri:2012} addresses this to some extent with measures on
the quality of a lineup, how numerically different, as best it can be
calculated, the data plot is from the null plots.

@@

\noindent The power of a lineup can therefore be estimated as the ratio of correct identifications $x$ out of $n$ viewings. (More details on these derivations can be found in \citet{majumder:2012}.)

%*** Under the null that the real is no different from the null, isn't $\pi_i=1/m$? Fixing this then I think the next paragraph on signal strength flows right from this one. It is when we are calculating signal strength that we start counting the number of observers who pick it.
%
%When there are multiple observers the significance level also depends on the number of independent observers ($n$). 
%Note that under the null hypothesis - i.e. when the real data is not different from the others - we expect $\pi_i = 1/m$.
%
%
%%The power of a test is the probability to reject the null hypothesis, when it is not true. In the framework of lineups, power is, effectively, the probability that the data plot is identified. , i.e. power can be estimated as $x/n$, where $x$ is the number of observers who correctly identifies the plot of the real data out of $n$ observers.
%
%%Let $x$ be the number of observers who
%%correctly identify the plot of the real data. The probability that one
%%observer correctly identifies the real data plot, in a size $m$
%%lineup, by chance is $1/m$. 
%
%\paragraph{Signal Strength}
%To incorporate the results from multiple
%observers think about the lineup as $m-1$ head-to-head competitions of
%the data plot against $m-1$ null plots, which the data plot has to
%win, to be picked as the plot with the most distinct features. 
%Assuming that observers are picking the plot with the strongest signal,  the
%probability that the data plot is picked by $x$ observers follows a
%Beta distribution for the maximum \citet{****}, which we call {\it signal strength} $\Xi$ \citet{majumder:2012}:
%\[
%\Xi =  \left( \frac{x}{n}\right)^{1/(m-1)}.
%\]
%Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data. $1-\Xi$ is analogous to the $p$-value of a classical hypothesis test. 
%
%
%
%
%*** The above paragraph needs more explanation, and a reference for the beta distribution. I don't think the Majumder paper really is the source for this result is it? It leads up to it, but this is the new work that is not submitted yet.
%
%*** Also need to refer to Niladri's ``Where's Waldo paper'' here too.
%
%%The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot: this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1]. Note, that the resulting interval is often too large to be useful.
%
%%We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features. Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum: this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
%
%%Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal. In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2011} that the signal strength is strongly correlated to the $p$-value of the corresponding statistical test for $\theta$.
%

>>>>

% !TEX root = comparison.tex
%% \section{Introduction} %for journal use above \firstsection{..} instead
%We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
So, you've just spent countless hours burning the late night candle
wax and discovered several astonishing features in the data. Your
exploratory graphics are quick to produce, interactive, flexible, and
fabulous for discovery, but now you need to present the information to
your boss, the co-workers, a class of students, or get it published in
that top-rated journal. Those exploratory graphics are unlikely to
communicate the discovered information to your audience efficiently,
or elegantly.  To decide on the best way to present the specific
information there can be many types of plots to choose from. The
decision process is assisted by past experience, personal aesthetic
preferences, the characteristics of the audience, and a good working
knowledge of general perceptual strengths and weaknesses of particular
displays \cite{cleveland:1984,heer:2010}. Perceptual studies are
thinly spread for data visualization and will probably
only cover very broad perceptual principles, primarily because they
tend to be based solely on simulated data. In order to study
perceptual principles in a controlled experimental setting simulated
data is unavoidable. This provides principles that may not apply
closely enough to a specific analysis task.  Faced with design
decisions for a very specific task requires a closer level of
detail. Focus groups \cite{Mazza07} can find small problems with
designs and move us closer to a final handful of possible
designs. Usability studies and case studies
(e.g. \cite{Plaisant:2004}, \cite{Shneiderman:2006}) typically focus
on the use of particular software for solving data exploration tasks.
These approaches lack broad audience validity and
objectiveness. Several papers indicate continuing issues arising with,
and suggestions for improving, evaluation of information
visualizations
\cite{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}.

@@

Here we suggest a new approach using recent research on visual inference \cite{buja:2009,wickham:2010}. In visual inference, data plots are considered to be test statistics, and these are compared with plots of data generated from a null hypothesis using a lineup. The null hypothesis underlying a lineup is that there is no real structure visible in the data plot, that any patterns seen are consistent with randomness, or from a known model. If the null hypothesis is true then the plot of the data will not be distinguishable form the plots of null data. A lineup is similar to the police lineup, and this is where the name comes from. The suspect (the plot of the data) is placed randomly in a field of plots of null data, and an impartial viewer is asked to identify the plot that is the most different. If the viewer identifies the suspect it lends statistical significance to the conclusion that the data is not consistent with the null data. Figure \ref{lineup} has an example of a lineup. Overlaid density plots are used to display the distribution of two groups (blue, red) of data. We are interested to know whether the data has a significant difference between the centers of the two groups. Plot numbered 13 contains the data. Did you pick that one? In this plot the blue group has a flat density whereas the density of red group is more peaked and concentrated to the left of the blue.

@@

The previous work \cite{buja:2009,wickham:2010} describes the use of lineups for evaluating discoveries made in exploratory analysis. Here we describe the use of lineups for evaluating plot design for communicating findings visually. Different lineups are created, using the same data, same null plots and positions within the lineup, but different plot designs. Comparing how long and how accurately the viewers take to report the feature of interest will assess which design is better for the task. To investigate the benefit of lineups for plot design evaluation we have conducted an experiment that also contains simulated data, where the data is constructed to match and expand on the features of interest in the data. For example, in a simple situation, where the discovered feature of interest is a difference between the centers of two distributions, simulated data would be constructed roughly matching the distribution of the data that varies the distance between centers, and also spread and sample size. Lineups of these simulated data (embedded with simulated null data plots) are generated and evaluated along with the lineups of data. The simulation study provides a backdrop for the real data, which enables the broad applicability to be studied and a gauge for the strength of the pattern in the data.

@@

Lineups provide a way to statistically quantify the significance of the finding \cite{majumder:2011}: think of the lineup as a set of $m$ plots, one of which is the data. (In Figure \ref{lineup}, $m=20$.)  The probability that an observer picks the data plot just by chance, i.e. when it is really *not* different from the other null plots, is $1/m$.  If an observer is able to identify the data plot from the lineup, we reject the null hypothesis.  This sets our Type I error rate, $\alpha$, at a level of $1/m$.

@@

When comparing different tests of the same quantity, we consider that
test better if it has greater {\it power}.  The power of a test is the
probability to reject the null hypothesis, irrespective of whether it
is true or false -- in a lineup this is the probability that an
observer identifies the plot of the real data.  Analytically, power is
usually difficult to calculate because it is requires specific use of
an alternative hypothesis to calculate the probability.  Work in
\cite{niladri:2012} addresses this to some extent with measures on
the quality of a lineup, how numerically different, as best it can be
calculated, the data plot is from the null plots.

@@

\noindent The power of a lineup can therefore be estimated as the ratio of correct identifications $x$ out of $n$ viewings. (More details on these derivations can be found in \cite{majumder:2012}.)

%*** Under the null that the real is no different from the null, isn't $\pi_i=1/m$? Fixing this then I think the next paragraph on signal strength flows right from this one. It is when we are calculating signal strength that we start counting the number of observers who pick it.
%
%When there are multiple observers the significance level also depends on the number of independent observers ($n$). 
%Note that under the null hypothesis - i.e. when the real data is not different from the others - we expect $\pi_i = 1/m$.
%
%
%%The power of a test is the probability to reject the null hypothesis, when it is not true. In the framework of lineups, power is, effectively, the probability that the data plot is identified. , i.e. power can be estimated as $x/n$, where $x$ is the number of observers who correctly identifies the plot of the real data out of $n$ observers.
%
%%Let $x$ be the number of observers who
%%correctly identify the plot of the real data. The probability that one
%%observer correctly identifies the real data plot, in a size $m$
%%lineup, by chance is $1/m$. 
%
%\paragraph{Signal Strength}
%To incorporate the results from multiple
%observers think about the lineup as $m-1$ head-to-head competitions of
%the data plot against $m-1$ null plots, which the data plot has to
%win, to be picked as the plot with the most distinct features. 
%Assuming that observers are picking the plot with the strongest signal,  the
%probability that the data plot is picked by $x$ observers follows a
%Beta distribution for the maximum \cite{****}, which we call {\it signal strength} $\Xi$ \cite{majumder:2012}:
%\[
%\Xi =  \left( \frac{x}{n}\right)^{1/(m-1)}.
%\]
%Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data. $1-\Xi$ is analogous to the $p$-value of a classical hypothesis test. 
%
%
%
%
%*** The above paragraph needs more explanation, and a reference for the beta distribution. I don't think the Majumder paper really is the source for this result is it? It leads up to it, but this is the new work that is not submitted yet.
%
%*** Also need to refer to Niladri's ``Where's Waldo paper'' here too.
%
%%The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot: this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1]. Note, that the resulting interval is often too large to be useful.
%
%%We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features. Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum: this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
%
%%Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal. In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2011} that the signal strength is strongly correlated to the $p$-value of the corresponding statistical test for $\theta$.
%
